---
title: "Lab 3: Multi-Source Retail Sales Data Integration & Analysis"
author: "Siddharth Metkari"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction

This lab focuses on integrating retail sales data from multiple
file formats and performing data cleaning, analysis, customer
segmentation, and SQL-based analysis.

The analysis uses CSV, JSON, Excel, and SQLite data sources.

The major objectives are:

- Import and clean data from multiple sources
- Integrate transactions, products, and customer information
- Analyze sales, products, countries, and customers
- Classify customers based on purchase value
- Store the integrated data in SQLite
- Execute SQL queries
- Generate business insights

# Setup

## Load Required Packages

```{r setup, message=FALSE, warning=FALSE}

library(tidyverse)
library(jsonlite)
library(readxl)
library(writexl)
library(DBI)
library(RSQLite)

```

## Set Working Directory

The notebook is designed to work from the `Lab3_Retail_Analysis`
folder containing the Excel dataset and generated source files.

```{r working-directory}

if (
  !file.exists("Online Retail.xlsx") &&
  dir.exists("Lab3_Retail_Analysis")
) {
  setwd("Lab3_Retail_Analysis")
}

getwd()

list.files()

```

```{r file-check}

if (!file.exists("Online Retail.xlsx")) {
  stop(
    paste(
      "Online Retail.xlsx was not found.",
      "Make sure the file is inside the Lab3_Retail_Analysis folder."
    )
  )
}

```

# Task 1: Import and Clean the Data

## Load the Original Dataset

The original Online Retail dataset is stored in Excel format.

```{r load-original}

retail <- read_excel("Online Retail.xlsx")

```

## Dataset Dimensions

```{r original-dimensions}

dim(retail)

```

## First Six Records

```{r original-head}

head(retail)

```

## Dataset Structure

```{r original-structure}

str(retail)

```

## Create the Transactions CSV Source

The transaction source contains invoice, product, customer,
quantity, and transaction date information.

```{r create-transactions}

transactions_source <- retail %>%
  select(
    InvoiceNo,
    StockCode,
    CustomerID,
    Quantity,
    InvoiceDate
  )

write_csv(
  transactions_source,
  "transactions.csv"
)

```

## Create the Products JSON Source

The product source contains product codes, descriptions, and
unit prices.

A single record is maintained for each `StockCode` to avoid
duplicating transaction records during the integration step.

```{r create-products}

products_source <- retail %>%
  select(
    StockCode,
    Description,
    UnitPrice
  ) %>%
  filter(
    !is.na(StockCode)
  ) %>%
  group_by(
    StockCode
  ) %>%
  summarise(
    Description =
      if (
        all(is.na(Description))
      ) {
        NA_character_
      } else {
        first(na.omit(Description))
      },
    UnitPrice =
      if (
        all(is.na(UnitPrice))
      ) {
        NA_real_
      } else {
        median(
          UnitPrice,
          na.rm = TRUE
        )
      },
    .groups = "drop"
  )

write_json(
  products_source,
  "products.json",
  pretty = TRUE,
  auto_unbox = TRUE
)

```

## Create the Customers Excel Source

The customer source contains customer IDs and countries.

```{r create-customers}

customers_source <- retail %>%
  select(
    CustomerID,
    Country
  ) %>%
  filter(
    !is.na(CustomerID)
  ) %>%
  group_by(
    CustomerID
  ) %>%
  summarise(
    Country =
      if (
        all(is.na(Country))
      ) {
        NA_character_
      } else {
        first(na.omit(Country))
      },
    .groups = "drop"
  )

write_xlsx(
  customers_source,
  "customers.xlsx"
)

```

## Verify the Three Source Files

```{r verify-source-files}

source_files <- c(
  "transactions.csv",
  "products.json",
  "customers.xlsx"
)

source_check <- data.frame(
  File = source_files,
  Exists = file.exists(source_files)
)

source_check

```

## Import CSV

```{r import-csv}

transactions <- read_csv(
  "transactions.csv",
  show_col_types = FALSE
)

head(transactions)

```

## Import JSON

```{r import-json}

products <- fromJSON(
  "products.json"
)

head(products)

```

## Import Excel

```{r import-customers}

customers <- read_excel(
  "customers.xlsx"
)

head(customers)

```

## Dataset Dimensions

```{r source-dimensions}

dim(transactions)

dim(products)

dim(customers)

```

## Check Missing Values

Missing values are checked before cleaning so that problematic
records can be identified.

```{r missing-values}

transactions_missing <- colSums(
  is.na(transactions)
)

products_missing <- colSums(
  is.na(products)
)

customers_missing <- colSums(
  is.na(customers)
)

transactions_missing
products_missing
customers_missing

```

## Check Duplicate Records

```{r duplicates}

transaction_duplicates <- sum(
  duplicated(transactions)
)

product_duplicates <- sum(
  duplicated(products)
)

customer_duplicates <- sum(
  duplicated(customers)
)

transaction_duplicates
product_duplicates
customer_duplicates

```

## Remove Duplicate Records

Duplicate transactions are removed to avoid double counting.

Product and customer datasets are reduced to one record per
identifier.

```{r remove-duplicates}

transactions <- transactions %>%
  distinct()

products <- products %>%
  distinct(
    StockCode,
    .keep_all = TRUE
  )

customers <- customers %>%
  distinct(
    CustomerID,
    .keep_all = TRUE
  )

```

## Remove Transactions with Missing Customer IDs

Customer-level analysis requires valid customer identifiers.

```{r remove-missing-customers}

missing_customer_count <- sum(
  is.na(transactions$CustomerID)
)

missing_customer_count

transactions <- transactions %>%
  filter(
    !is.na(CustomerID)
  )

```

## Remove Invalid Quantities

Transactions with zero or negative quantities are removed because
the analysis focuses on positive sales transactions.

```{r clean-quantity}

invalid_quantity_count <- sum(
  transactions$Quantity <= 0,
  na.rm = TRUE
)

invalid_quantity_count

transactions <- transactions %>%
  filter(
    Quantity > 0
  )

```

## Remove Invalid Prices

Products with missing or non-positive prices are removed.

```{r clean-price}

invalid_price_count <- sum(
  products$UnitPrice <= 0,
  na.rm = TRUE
)

invalid_price_count

products <- products %>%
  filter(
    !is.na(UnitPrice),
    UnitPrice > 0
  )

```

## Final Clean Dataset Dimensions

```{r final-clean-dimensions}

dim(transactions)

dim(products)

dim(customers)

```

## Cleaning Interpretation

The following cleaning decisions were applied:

1. Duplicate records were removed to prevent double counting.
2. Transactions with missing `CustomerID` were removed because
   customer-level analysis requires valid customer identifiers.
3. Transactions with `Quantity <= 0` were removed because the
   analysis focuses on positive sales.
4. Products with missing or non-positive `UnitPrice` were removed.
5. Missing values were inspected before integration.
6. Product records were reduced to one record per `StockCode`
   to prevent transaction duplication during joins.

# Task 2: Integrate Multiple Data Sources

## Check Unmatched Products

The transaction dataset is checked against the product dataset
using `StockCode`.

```{r unmatched-products}

unmatched_products <- transactions %>%
  anti_join(
    products,
    by = "StockCode"
  )

nrow(unmatched_products)

```

## Check Unmatched Customers

The transaction dataset is checked against the customer dataset
using `CustomerID`.

```{r unmatched-customers}

unmatched_customers <- transactions %>%
  anti_join(
    customers,
    by = "CustomerID"
  )

nrow(unmatched_customers)

```

## Join Transactions with Products

A left join is used so that valid transaction records are retained
while product information is matched where available.

```{r join-products}

sales_data <- transactions %>%
  left_join(
    products,
    by = "StockCode"
  )

```

## Join Customer Information

Customer information is then added using `CustomerID`.

```{r join-customers}

sales_data <- sales_data %>%
  left_join(
    customers,
    by = "CustomerID"
  )

```

## Calculate Revenue

Revenue is calculated using:

**Revenue = Quantity × UnitPrice**

```{r calculate-revenue}

sales_data <- sales_data %>%
  mutate(
    Revenue = Quantity * UnitPrice
  )

```

## Verify the Integrated Dataset

```{r integrated-dataset}

dim(sales_data)

head(
  sales_data,
  10
)

```

## Missing Values After Integration

```{r integrated-missing}

colSums(
  is.na(sales_data)
)

```

## Join Interpretation

A `left_join()` was selected because all valid transaction records
should be retained while product and customer information is
matched wherever available.

This approach also allows unmatched records to be identified
without unnecessarily removing valid transactions.

# Task 3: Sales and Customer Analysis

## Total Sales Revenue

```{r total-revenue}

total_revenue <- sales_data %>%
  summarise(
    Total_Revenue =
      sum(
        Revenue,
        na.rm = TRUE
      )
  )

total_revenue

```

### Interpretation

The total revenue represents the overall sales value generated
after the data cleaning and integration process.

## Top 5 Products by Revenue

```{r top-products}

top_5_products <- sales_data %>%
  group_by(
    StockCode,
    Description
  ) %>%
  summarise(
    Revenue =
      sum(
        Revenue,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) %>%
  arrange(
    desc(Revenue)
  ) %>%
  slice_head(
    n = 5
  )

top_5_products

```

### Interpretation

The products with the highest revenue are important contributors
to overall sales and can be considered for inventory and marketing
strategies.

## Top 5 Countries by Revenue

```{r top-countries}

top_5_countries <- sales_data %>%
  group_by(
    Country
  ) %>%
  summarise(
    Revenue =
      sum(
        Revenue,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) %>%
  arrange(
    desc(Revenue)
  ) %>%
  slice_head(
    n = 5
  )

top_5_countries

```

### Interpretation

The countries with the highest revenue represent the strongest
markets in the dataset.

## Top 5 Customers by Purchase Value

```{r top-customers}

top_5_customers <- sales_data %>%
  group_by(
    CustomerID
  ) %>%
  summarise(
    Total_Purchase_Value =
      sum(
        Revenue,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) %>%
  arrange(
    desc(Total_Purchase_Value)
  ) %>%
  slice_head(
    n = 5
  )

top_5_customers

```

### Interpretation

The top customers contribute a significant portion of the
company's revenue and may be suitable targets for loyalty and
retention strategies.

## Customer Value Classification

Customer purchase values are calculated first.

```{r customer-value}

customer_value <- sales_data %>%
  group_by(
    CustomerID
  ) %>%
  summarise(
    Total_Purchase_Value =
      sum(
        Revenue,
        na.rm = TRUE
      ),
    .groups = "drop"
  )

head(customer_value)

```

## Calculate Quartile Thresholds

```{r customer-quartiles}

q1 <- quantile(
  customer_value$Total_Purchase_Value,
  0.25,
  na.rm = TRUE
)

q2 <- quantile(
  customer_value$Total_Purchase_Value,
  0.50,
  na.rm = TRUE
)

q3 <- quantile(
  customer_value$Total_Purchase_Value,
  0.75,
  na.rm = TRUE
)

q1
q2
q3

```

## Classify Customers Using `case_when()`

Customers are classified into four groups:

- Low Value
- Medium Value
- High Value
- Premium

```{r customer-classification}

customer_value <- customer_value %>%
  mutate(
    Customer_Category = case_when(
      Total_Purchase_Value <= q1 ~ "Low Value",
      Total_Purchase_Value <= q2 ~ "Medium Value",
      Total_Purchase_Value <= q3 ~ "High Value",
      TRUE ~ "Premium"
    )
  )

customer_value

```

## Customer Category Counts

```{r customer-category-counts}

customer_category_counts <- customer_value %>%
  count(
    Customer_Category
  )

customer_category_counts

```

### Interpretation

Customer classification helps identify differences in customer
value and can support targeted retention and marketing strategies.

## Market Performance

```{r market-performance}

market_performance <- sales_data %>%
  group_by(
    Country
  ) %>%
  summarise(
    Revenue =
      sum(
        Revenue,
        na.rm = TRUE
      ),
    .groups = "drop"
  ) %>%
  filter(
    !is.na(Country)
  ) %>%
  filter(
    Revenue > 0
  ) %>%
  arrange(
    desc(Revenue)
  )

head(
  market_performance,
  10
)

```

## High-Performing Market

```{r high-market}

high_performing_market <- market_performance %>%
  slice_head(
    n = 1
  )

high_performing_market

```

## Underperforming Market

```{r low-market}

underperforming_market <- market_performance %>%
  slice_tail(
    n = 1
  )

underperforming_market

```

### Market Interpretation

```{r market-interpretation, results='asis'}

cat(
  "The high-performing market is ",
  high_performing_market$Country,
  " with revenue of ",
  round(high_performing_market$Revenue, 2),
  "."
)

cat("\n\n")

cat(
  "The underperforming market is ",
  underperforming_market$Country,
  " with revenue of ",
  round(underperforming_market$Revenue, 2),
  "."
)

```

# Task 4: SQLite Database and SQL Queries

## Create SQLite Database

```{r sqlite-connect}

con <- dbConnect(
  RSQLite::SQLite(),
  "retail_sales.db"
)

```

## Store Integrated Dataset

The final integrated dataset is stored in a SQLite table named
`retail_sales`.

```{r sqlite-write}

dbWriteTable(
  con,
  "retail_sales",
  sales_data,
  overwrite = TRUE
)

```

## Verify Database Tables

```{r sqlite-tables}

dbListTables(con)

```

## Verify Table Fields

```{r sqlite-fields}

dbListFields(
  con,
  "retail_sales"
)

```

## SQL Query 1: Top 5 Customers

The first SQL query identifies the five customers with the
highest total revenue.

```{r sql-query-1}

sql_query_1 <- "

SELECT
    CustomerID,
    SUM(Revenue) AS TotalRevenue
FROM retail_sales
GROUP BY CustomerID
ORDER BY TotalRevenue DESC
LIMIT 5;

"

top_5_customers_sql <- dbGetQuery(
  con,
  sql_query_1
)

top_5_customers_sql

```

### Interpretation

This query identifies the most valuable customers based on their
total purchase value.

## SQL Query 2: Revenue by Country

The second SQL query calculates total revenue for each country.

```{r sql-query-2}

sql_query_2 <- "

SELECT
    Country,
    SUM(Revenue) AS TotalRevenue
FROM retail_sales
GROUP BY Country
ORDER BY TotalRevenue DESC;

"

revenue_by_country_sql <- dbGetQuery(
  con,
  sql_query_2
)

head(
  revenue_by_country_sql,
  10
)

```

### Interpretation

This query allows the business to compare revenue performance
across different countries and identify strong and weak markets.

## Database Record Count

```{r database-count}

database_count <- dbGetQuery(
  con,
  "
  SELECT COUNT(*) AS TotalRecords
  FROM retail_sales;
  "
)

database_count

```

## Close Database Connection

```{r close-database}

dbDisconnect(
  con
)

```

# Business Insights

## Insight 1: Overall Revenue

```{r insight-revenue, results='asis'}

cat(
  "The cleaned and integrated dataset generated total revenue of ",
  round(total_revenue$Total_Revenue, 2),
  "."
)

```

This provides an overall measure of the sales value represented
by the cleaned dataset.

## Insight 2: Highest-Revenue Product

```{r insight-product, results='asis'}

cat(
  "The highest-revenue product is '",
  top_5_products$Description[1],
  "' with revenue of ",
  round(top_5_products$Revenue[1], 2),
  "."
)

```

This product represents an important contributor to overall
sales and can be considered for inventory and marketing decisions.

## Insight 3: Market Performance

```{r insight-market, results='asis'}

cat(
  high_performing_market$Country,
  " is the highest-performing market with revenue of ",
  round(high_performing_market$Revenue, 2),
  ", while ",
  underperforming_market$Country,
  " has the lowest revenue among the analyzed markets."
)

```

This indicates that sales performance differs significantly
across countries and that weaker markets may require additional
marketing or market-development strategies.

# Final Output File Verification

```{r final-files}

output_files <- c(
  "transactions.csv",
  "products.json",
  "customers.xlsx",
  "retail_sales.db"
)

final_check <- data.frame(
  File = output_files,
  Exists = file.exists(output_files)
)

final_check

```

# Conclusion

This lab demonstrated a complete retail data integration and
analytics workflow.

Data was imported from CSV, JSON, and Excel sources, cleaned,
validated, and integrated using joins.

The integrated dataset was then used to calculate total revenue,
identify top products, countries, and customers, classify customers
according to purchase value, and evaluate market performance.

Finally, the integrated dataset was stored in SQLite and analyzed
using SQL queries.

The resulting analysis provides useful information for understanding
customer value, product performance, and market performance.

getwd()

rmarkdown::render("Lab3_Retail_notebook.Rmd")
setwd("Lab3_Retail_Analysis")
